Optimizing future biodiversity sampling by citizen scientists

### 2019-04-25 paper/main.tex
changes from Jodi, RM, and some from Will

\title{Optimizing the future of biodiversity sampling by citizen scientists}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,b,1]{Corey T. Callaghan}
\author[c,]{Alistair G. B. Poore}
\author[b,a]{Richard E. Major}
\author[b,a]{Jodi J. L. Rowley}
\author[a,c]{William K. Cornwell}

\affil[a]{Centre for Ecosystem Science, School of Biological, Earth and Environmental Sciences, UNSW Sydney, Sydney, 2052, NSW, Australia}
\affil[b]{Australian Museum Research Institute, Australian Museum, Sydney, 2000, NSW, Australia}
\affil[c]{Ecology and Evolution Research Centre, School of Biological, Earth and Environmental Sciences, UNSW Sydney, Sydney, 2052, NSW, Australia}

@@

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Citizen science has the potential to significantly contribute to ecology, conservation, and natural resource management in the future. But we need to better understand how to maximize the value from the collective effort of Citizen scientists. We provide a framework, specific to an intended statistical outcome (e.g., population trend model), relying on statistical leverage to assign the marginal value of a given citizen science observation. This framework is dynamic, and can be calculated for a given day, providing a map of expected value for a given citizen science observation. We recommend citizen scientists, and citizen science project managers, adapt this framework to their needs, with a focus on maximizing the value of citizen science efforts for biodiversity management.}

@@

\begin{abstract}
We are currently in the midst of the sixth great extinction, and understanding biodiversity trends in space and time is essential for curbing biodiversity loss and prioritizing limited resources for conservation. Concomitantly, funding for ecology and conservation research is disappearing. In turn, scientists are increasingly using citizen science data to extract information about population trends, with the aim of providing management recommendations. But these data are generally 'noisy' with a suite of redundancies and gaps arising from the associated structure of the project. Here, we ask whether these data can be improved --- maximizing the information content extracted from citizen science datasets --- by optimizing sampling in space and time by citizen scientists. We provide a novel framework, based on statistical leverage, which assigns every citizen science observation a marginal value derived from the importance of an observation to overall population trends. We test how a suite of dynamic characteristics of sampling in space and time influence the marginal value of citizen science sampling events. We find that the `history' of a grid influences the future optimal sampling: the marginal value of a sampling event was most associated with the number of unique sampling days of a grid. We also find that grids should aim to decrease their median sampling interval. Our framework is robust to spatial-scale, easily adapted within a national park, entire state, or country. We conclude by using our parameterized model to produce dynamic maps, in space and time, predicting the expected marginal value in space for every day in 2018. Ultimately, we envision an approach where citizen scientists are guided and incentivized to sample sites with the highest expected marginal value.
\end{abstract}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{A}ssessing biodiversity trends in space and time is essential for conservation \cite{harrison2014assessing, wilson2011modelling, mcmahon2011improving, honrado2016fostering, yoccoz2001monitoring} \parshape=0. Reliable biodiversity trend estimates, at multiple spatial scales \cite{soberon2007assessing}, allow us to track our global progress in curbing biodiversity loss \cite{harrison2014assessing}. Unsurprisingly, reliable trend estimates are best derived from long-term \cite{lindenmayer2012value, magurran2010long}, well-designed surveys, in space and time \cite{harrison2014assessing, vellend2017estimates, kery2009trend}. But scientific funding for long-term ecological and conservation research is failing to keep pace with conservation research efforts and needs \cite{bakker2010changing}. Increasingly, government agencies, scientific researchers, and conservationists are turning to citizen science data to help inform the state of biodiversity at local \cite{callaghan2015efficacy, theobald2015global, sullivan2017using, loss2015linking}, regional \cite{barlow2015citizen, fox2011new}, and global scales \cite{chandler2017contribution, pocock2018vision, cooper2014invisible}.

Citizen science --- the cooperation between a range of experts and non-experts  --- is an incredibly diverse field \cite{jordan2015citizen}. Projects generally fall along a continuum based on the level of associated structure \cite{kelling2019using, welvaert2016citizen}, ranging from unstructured (e.g., opportunistic or incidental projects with little to no training required; iNaturalist) to structured (e.g., projects with specific objectives, rigorous protocols, and survey design; U.K. Butterfly Monitoring Scheme). The level of structure, in turn, influences the redundancies and gaps in the data, as well as the data quality of a particular project. For instance, observer skill \cite{kelling2015can}, time-of-day, number of participants in a group, and technological capabilities of a participant may influence the data collected by some, but not necessarily all, citizen science projects. Generalizeable among citizen science projects, however, are various redundancies and gaps (i.e., spatial and temporal biases) \cite{boakes2010distorted, bird2014statistical}. Observers submitting observations on weekends \cite{courter2013weekend}, sampling near roads and human settlements \cite{kelling2015taking}, and observers visiting known 'hotspots' for biodiversity \cite{geldmann2016determines} are all examples of such data constraints. These biases are not neccesarily restricted to citizen science projects, as our historical understanding of biodiversity in space and time is also biased; evidenced by natural history collections \cite{pyke2010biological, boakes2010distorted}. Indeed, many sampling methods have been proposed to optimally sample biodiversity by professional scientists \cite{etienne2005new, moreno2000assessing, colwell1994estimating, longino1997biodiversity, ferrarini2012biodiversity}, frequently dependent on spatial scale \cite{chase2013scale}. Structured citizen science projects generally adapt some form of optimal sampling (e.g., stratified sampling), but little attention has been given to optimal sampling by citizen scientists in unstructured citizen science projects \cite{harrison2014assessing}.

This is probably, in part, because these redundancies and gaps in the data are seen as a 'necessary hurdle', associated with such data \cite{parrish2018exposing}. Also, in the case of broad-scale biodiversity data collected at large voluminous scales (e.g., eBird), these various biases can generally be accounted for statistically \cite{isaac2014statistics, robinson2018correcting}, for instance, by filtering or subsetting data \cite{wiggins2011conservation}, pooling multiple data sources \cite{fithian2015bias}, or machine learning and hierarchical clustering techniques \cite{hochachka2012data, kelling2015taking}. Indeed, despite known biases, citizen science data have increased our knowledge of species distribution models \cite{bradsworth2017species, van2013opportunistic}, niche breadth \cite{tiago2017using}, biodiversity measurements \cite{stuart2017assessing, pocock2018vision}, phenological research \cite{la2014role, supp2015citizen}, invasive species detection \cite{pocock2017citizen, grason2018citizen}, and phylogeographical research \cite{bahls2014new, drury2019continent}. Still, estimating trends with citizen science data is best done with data from structured projects (i.e., less biases to account for) \cite{fox2011new}. But unstructured and semistructured projects are increasingly harnessed for trend detection \cite{walker2017using, kery2009trend, kery2010site, horns2018using, van2013occupancy, pagel2014quantifying}. From a conservation perspective, the goal is relatively straightforward: provide robust measures of species' trends through time, a critical component of the IUCN Red List Index \cite{baillie2008toward}. The robustness of these trend estimates is critical, and the goal should be to continuously decrease the standard error of trend estimates (e.g., Fig. 1). This is possible with citizen science data \cite{kery2010site, horns2018using, van2013occupancy, pagel2014quantifying} and the standard error is generally proportional to the number of observations, and appropriate sampling, through time (e.g.,\href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/Noisy_miner_gif.gif}{Video S1}).

\begin{figure}[!hb]
\centering
\includegraphics[width=.8\linewidth]{glm_examples_edited.png}
\caption{The ultimate goal in understanding population trends is to minimize the standard error for a population trend model, providing more robust measures of population trends. Shown here are four example population trend models, based on eBird data between 2010-2018, for Noisy Miner (top left), Hardhead (top right), Masked Lapwing (bottom left), and Crested Pigeon (bottom right).}
\label{fig1}
\end{figure}

The number of citizen science projects which are focused on ecological and environmental monitoring is increasing \cite{pocock2017diversity, theobald2015global}, highlighting the potential that citizen science holds for the future of ecology, conservation, and natural resource management \cite{pocock2018vision, silvertown2009new, soroye2018opportunistic, mckinley2017citizen}. But a major obstacle in the future use of citizen science data remains understanding how to best extract information from 'noisy' citizen science datasets \cite{parrish2018exposing}. As mentioned, this noise from citizen science \cite{bird2014statistical} can sometimes be alleviated using 'big-data' statistical approaches \cite{kelling2015taking}, but this is most applicable for data originating from large, successful citizen science projects --- with lots of data. What about projects that are just beginning? Or projects focusing on taxa that are less popular with the general public \cite{mair2016explaining, ward2014understanding}? Are there optimal strategies for sampling in space and time for estimating biodiversity trends?

Here, we investigate these questions with a specific objective: assess how spatial and temporal sampling influences trend detection of biodiversity. Our approach is dynamic: we are interested in the parameters that influence the marginal value of a given citizen science observation in both time and space. First, we summarize the parameters which we hyptohesized would influence the marginal value of a citizen science observation.

@@

\begin{itemize}
  \item \textbf{Whether a site was sampled}: if a site had been previously sampled or not. We predicted that unsampled sites would be marginally more valuable than sampled previously sampled sites.
  \item \textbf{Median sampling interval}: the median of the distribution of waiting times between samples at a site. We predicted that the median sampling interval would be positively associated with the value of a citizen science observation. I.e., observations from sites with high median waiting times would be more valuable than observations from sites with low median waiting times.
  \item \textbf{Days since last sample}: the number of days between samples at a site. We predicted that the days since the last sample would positively associate with the value of a citizen science observation.
  \item \textbf{Distance to the nearest sampled site}: the distance between the site in question and the nearest sampled site. We hypothesized that for the distance to the nearest sampled site would be positively associated with the value of a citizen science observation.
  \item \textbf{Nearest neighbor sampling interval}: the median sampling interval of the nearest neighbor. We hypothesized that this would be positively influence the value of a observation, whereby well-sampled areas (i.e., multiple sites near each other with low median sampling intervals) would have lower value observations.
  \item \textbf{Number of unique days sampled}: the total number of unique days sampled for a given site. We predicted that the total number of unique days would be positively associated with the value of an observation, whereby sites with lots of observations would receive value given the long-term data originating from them.
\end{itemize}

We test our predictions using a popular citizen science project --- eBird \cite{sullivan2009ebird} --- and > 5 million bird observations from the Greater Sydney Region, NSW, Australia. We fit presence/absence population trend models to 235 species throughout this region, based on data from 2010 to 2018 (Fig. 1), and calculate statistical leverage for every citizen science sampling event, for each species. The leverage is a measure of how 'valuable' a given observation is to the population trend model for each species. Each sampling event is then assigned a cumulative measure of leverage --- i.e., marginal value. We then test our predictions above using linear models, assessing the influence each of the parameters has on the trend detection of biodiversity. Finally, we use the parameterized model results to predict the expected leverage (i.e., marginal value) for a given grid in space and time, producing a dynamic map of expected marginal value of citizen science sampling for every day in 2018.

\section*{Results and Discussion}
To assess the robustness of our approach, we investigated our predictions at four different grid sizes: 5, 10, 25, and 50 km$^{2}$ grids. We found weak evidence that visiting an unsampled grid was marginally more valuable than visiting an already sampled grid. There was no statistical significance for the 5 km (p=0.669), and 10 km (p=0.093) grid sizes, but there was for the 25 km (p=0.035) grid size. At the 50 km grid size, this test was not possible because all grids had been sampled. These results suggests that stratified sampling --- an approach which aims for equal sampling among grids or sites --- \cite{baillie2008toward, longino1997biodiversity} is not the most applicable approach for detecting trends using citizen science data. In other words, citizen scientists contributing to eBird, likely a large number who are qualified naturalists in their own right \cite{callaghan2018unnatural}, are already sufficiently sampling biodiversity in space.

@@

For those sampled grids, however, we found significant relationships between a number of our predicted parameters and the marginal value of a sampling event (Fig. 2). In general, our parameters were positively associated with the marginal value of a sampling event. Of our parameters of interest, the number of unique days sampled (i.e., the `history' of a grid) had the strongest, positive, effect size, and this was robust to the grid size comparisons. The median sampling interval was also strongly associated with high marginal value samples, with an exception at the 25 km grid size. Distance to the nearest sampled grid and the nearest neighbor sampling interval influenced the marginal value of a sampling event less than the other parameters. Surprisingly, days since the last sample had less influence than the other parameters, albeit a positive association. The fact that days since last sample had a lower effect size than both the median sampling interval and the number of unique days sampled, likely highlights the value of `revisiting' a site in order to extract the maximum amount of information.

Together, these results suggest that there are possible improvements in biodiversity sampling for trend detection. In short, the `history' of a grid is important for considering whether to sample that grid: the number of unique days sampled was the strongest predictor, suggesting that observations from sites with a long time-series are marginally more valuable. Further, a secondary goal should be to decrease (i.e., left-shift the distribution) of the median sampling interval of the grids. The effect of number of days sampled was strongest at the largest grid size, suggesting that given all grids are sampled with some regularity --- as they are with larger grid sizes --- the importance of the `history' of a site is inflated.

We found generally consistent results, based on the grid size chosen, highlighting the robustness of spatial scale in our approach. It is critical to track biodiversity trends at multiple spatial scales \cite{soberon2007assessing}, as biodiversity estimates sometimes change dependent on the spatial scale \cite{chase2013scale}. Our approach of relying on statistical leverage as a measure of the value of a citizen science observation can be used regardless of whether a citizen science project is global, regional, or carried out in the constraints of a local park. In any instance, the area can be gridded --- the size of a grid, however, would be proportional to the spatial scale of the study --- and the framework implemented, updated dynamically.

Providing dynamic feedback to citizen science participants has proved successful for many citizen science projects \cite{rowley2019frogid, wiggins2011conservation, xue2016avicaching}. This feedback is generally in the form of leaderboards, presenting the number of submissions or number of unique species someone has contributed \cite{wood2011ebird}. We used our fitted models to predict the expected value of a given citizen science observation, dynamically, for any given day (e.g., Fig. 3, \href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/dynamic_map.gif}{Video S2}). This approach required us to look backward first, using a model with all observations for 2018, based on statistical leverage calculated from 2010-2018, in order to look forward and pedict the expected leverage for any given day. We envision a dynamic approach (Video S2) in the future of citizen science projects, which would ultimately guide participants to sites which \textit{should} be sampled on any given day --- or in a given week, month, or year. In this instance, leaderboards would move past numbers of species or submissions and complementary leaderboards could be derived based on a participant's cumulative value to the citizen science dataset, for biodiversity outcomes. Instead of participants preferrentially chasing specific species, this approach would guide participants to the sites with the highest expected marginal value. For example, we imagine visitor centers across the world at national parks or urban greenspaces providing their visitors on any given day a localized map showing which trail someone should visit if they are interested in contributing the greatest value to that park's biodiversity knowledge, through citizen science. The global pull of ecotourism \cite{sharpley2006ecotourism} is increasing exponentially, creating the potential for people to contribute to local biodiversity knowledge in areas that are traditionally undersampled, and with this framework, the collective effort of citizen scientists can be maximized.

\begin{figure}[!hb]
\centering
\includegraphics[width=.8\linewidth]{example_maps.png}
\caption{A map of predicted expected marginal value for six different days in 2018, throughout the Greater Sydney Region, showing the highest valued grid cells that would optimize the collective knowledge on biodiversity trends throughout the Greater Sydney Region. This is a dynamic approach, where the predictions are updated as new observations are submitted to the citizen science database, and we have done this for every day in 2018 (\href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/dynamic_map.gif}{Video S2}).}
\label{fig3}
\end{figure}

We focused our framework on a specific statistical outcome: trend detection. Many other ecological outcomes arise from citizen science datasets, including species distribution models \cite{bradsworth2017species, van2013opportunistic}, phylogeographical research \cite{bahls2014new, drury2019continent}, invasive species detection \cite{pocock2017citizen, grason2018citizen}, or phenological research \cite{la2014role, supp2015citizen}. Each potential outcome will have different optimal sampling strategies in space and time, likely with nuanced trade-offs between outcomes. But these different outcomes can still be quantified in the same framework we introduce here. For example, an intended outcome of a spcecies distribution model would likely place greater value on observations from unsampled grids \cite{crawley2001scale} than for species trend detection. Our framework is robust, with the necessary piece of information being a statistical model based on an intended ecological/management outcome.

\subsection*{Conclusions}
Since eBird's inception in 2002, citizen scientists have collectively contributed > 30 million effort hours. And this is only one citizen science project. Clearly, the cumulative effort put-forth by citizen scientists is immense, and arguably, citizen science will continue to shape the future of ecology and conservation --- as it has substantially for the past couple centuries \cite{silvertown2009new} --- with an increasingly critical role \cite{mckinley2017citizen, pocock2018vision} in monitoring of biodiversity. But we need to look towards the future. Are there mechanisms we can put in place now which will increase our collective knowledge gleaned from citizen science datasets for biodiversity in the future? We highlighted general rules which could help to guide citizen science participants to better sampling in space and time: the number of unique days sampled and the largest median sampling intervals both positively correlate with the marginal value of a citizen science observation. Moreover, we demonstrate a framework which practicioners can implement to better optimize their sampling designs, specific to a citizen science project and an intended goal.

>>>>

\title{Optimizing future biodiversity sampling by citizen scientists}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,b,1]{Corey T. Callaghan}
\author[c,]{Alistair G. B. Poore}
\author[b,a]{Richard E. Major}
\author[b,a]{Jodi J. L. Rowley}
\author[c,a]{William K. Cornwell}

\affil[a]{Centre for Ecosystem Science, School of Biological, Earth and Environmental Sciences, UNSW Sydney, Sydney, 2052, NSW, Australia}
\affil[b]{Australian Museum Research Institute, Australian Museum, Sydney, 2010, NSW, Australia}
\affil[c]{Ecology and Evolution Research Centre, School of Biological, Earth and Environmental Sciences, UNSW Sydney, Sydney, 2052, NSW, Australia}

@@

% Please add here a significance statement to explain the relevance of your work
\significancestatement{The challenge of monitoring biodiversity during the ongoing extinction crisis is unlikely to be matched by sufficient funding to professional scientists. Citizen science may fill much of the shortfall, but currently, citizen science sampling is haphazard, and we can do a better job maximizing both the marginal value of each person's effort and the total information content of a given dataset. We provide a dynamic framework to do this, showing where and when to sample biodiversity to maximize value per effort. We recommend that citizen science projects adapt this framework to their needs, with a focus on maximizing the marginal value of each citizen science contribution to inform biodiversity conservation goals.}

@@

\begin{abstract}
We are currently in the midst of earth's sixth extinction event, and measuring biodiversity trends in space and time is essential for prioritizing limited resources for conservation and curbing biodiversity loss. At the same time, the scope of the necessary monitoring is massively overwhelming the funding for professional scientific monitoring. In response, scientists are increasingly using citizen science data to estimate population trends, with the aim of providing prioritizing conservation efforts. But citizen science data are 'noisy', with redundancies and gaps arising from unstructured human behaviors in space and time. Here, we ask whether the information in these data can be improved --- maximizing the information content extracted from citizen science datasets for trend estimation. We develop and execute a novel framework, based on statistical leverage, which assigns every citizen science observation a marginal value derived from the importance of an observation to our understanding of overall population trends for a given geographic region. We then make this framework predictive, estimating the expected marginal value of future biodiversity observations. We find that past observations are useful in forecasting where high value observations will occur in the future. Interestingly, we find high marginal value in both `hotspots` which are frequently sampled locations and `coldspots` areas far from recent sampling, suggesting that an optimal sampling regime balances `hotspot` sampling with a spread across the landscape. Our framework is applicable at any spatial scale, easily adapted within a national park, entire country, or global. Because expected value is dependent on the behavior of all other citizen scientists, our model produces dynamic maps of expected marginal value through time. Ultimately, we envision an approach where citizen scientists are guided and incentivized to sample sites with the highest expected marginal value, collectively maximizing our understanding of biodiversity in space and time, and our ability to conserve it.
\end{abstract}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{A}ssessing biodiversity trends in space and time is essential for conservation \cite{harrison2014assessing, wilson2011modelling, mcmahon2011improving, honrado2016fostering, yoccoz2001monitoring} \parshape=0. Reliable biodiversity trend estimates, at multiple spatial scales \cite{soberon2007assessing}, allow us to track our global progress in curbing biodiversity loss while managing our scarce conservation resources \cite{harrison2014assessing}. Unsurprisingly, reliable trend estimates are best derived from long-term \cite{lindenmayer2012value, magurran2010long}, well-designed surveys, carried out over a wide spatial and temporal scale \cite{harrison2014assessing, vellend2017estimates, kery2009trend}. But scientific funding for long-term ecological and conservation research is failing to keep pace with conservation needs \cite{bakker2010changing, rios2018fieldwork}. Increasingly, government agencies, scientific researchers, and conservationists are turning to citizen science data to help inform the state of biodiversity at local \cite{callaghan2015efficacy, theobald2015global, sullivan2017using, loss2015linking}, regional \cite{barlow2015citizen, fox2011new}, and global scales \cite{chandler2017contribution, pocock2018vision, cooper2014invisible}.

Citizen science --- the cooperation between scientific experts and non-experts  --- is an incredibly diverse field \cite{jordan2015citizen}. Projects generally fall along a continuum based on the level of associated structure \cite{kelling2019using, welvaert2016citizen}, ranging from unstructured (e.g., opportunistic or incidental projects which require little to no training required; iNaturalist) to structured (e.g., projects with specific objectives, rigorous protocols, and survey design; U.K. Butterfly Monitoring Scheme). The level of structure, in turn, influences the degree of redundancies and gaps in the data, as well as the overall data quality of a particular project. For instance, observer skill \cite{kelling2015can}, number of participants in a group, and the technological capabilities of a participant may influence the data collected by some, but not necessarily all, citizen science projects. Generalizeable among citizen science projects, however, are various redundancies and gaps (i.e., spatial and temporal biases) \cite{boakes2010distorted, bird2014statistical}. Observers submitting observations on weekends \cite{courter2013weekend}, sampling near roads and human settlements \cite{kelling2015taking}, and visiting known 'hotspots' for biodiversity \cite{geldmann2016determines} are all examples of how unstructured human behavior leads to redundancies and gaps in citizen science data. These biases are not restricted to citizen science projects. Indeed, our historical understanding of biodiversity in space and time is also biased; evidenced by natural history collections \cite{pyke2010biological, boakes2010distorted} and many sampling methods have been proposed to optimize biodiversity sampling by professional scientists \cite{etienne2005new, moreno2000assessing, colwell1994estimating, longino1997biodiversity, ferrarini2012biodiversity}, frequently dependent on spatial scale \cite{chase2013scale}. While structured citizen science projects often adapt some aspects of optimal sampling in their methods (e.g., stratified sampling), little attention has been given to optimal sampling in unstructured citizen science projects \cite{harrison2014assessing}.

One of the reasons that optimal sampling has been largely ignored in unstructured citizen science projects is because redundancies and gaps in the data are seen as a "necessary hurdle" \cite{parrish2018exposing}. Also, in the case of broad-scale biodiversity data collected at large voluminous scales (e.g., eBird), the biases can generally be accounted for statistically \cite{isaac2014statistics, robinson2018correcting}, for instance, by filtering or subsetting data \cite{wiggins2011conservation}, pooling multiple data sources \cite{fithian2015bias}, or machine learning and hierarchical clustering techniques \cite{hochachka2012data, kelling2015taking}. Indeed, despite known biases, citizen science data have increased our knowledge of species distribution models \cite{bradsworth2017species, van2013opportunistic}, niche breadth \cite{tiago2017using}, biodiversity measurements \cite{stuart2017assessing, pocock2018vision}, phenological research \cite{la2014role, supp2015citizen}, invasive species detection \cite{pocock2017citizen, grason2018citizen}, and phylogeographical research \cite{bahls2014new, drury2019continent}.

Despite the potential ecological outcomes from citizen science data, estimating biodiversity trends is perhaps the most important, given the current need to efficiently and effectively monitor biodiversity \cite{harrison2014assessing, honrado2016fostering, yoccoz2001monitoring}. From a conservation perspective, the goal is relatively straightforward: provide robust measures of species' trends through time, a critical component of the IUCN Red List Index \cite{baillie2008toward}. Estimating trends with citizen science data is best done with data from structured projects (i.e., less biases to account for, generally resulting in less standard errors) \cite{fox2011new}. But unstructured and semistructured projects are increasingly harnessed for trend detection \cite{walker2017using, kery2009trend, kery2010site, horns2018using, van2013occupancy, pagel2014quantifying}. The robustness of these trend estimates is critical, and the goal should be to continuously decrease the standard error surrounding these estimates (e.g., Fig. 1). Unsurprisingly, the standard error is generally proportional to the number of observations, as well as appropriate sampling, through time (e.g.,\href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/Noisy_miner_gif.gif}{Video 1}).

\begin{figure}[!hb]
\centering
\includegraphics[width=.8\linewidth]{glm_examples_edited.png}
\caption{The ultimate goal in understanding population trends is to minimize the standard error for a population trend model, providing more robust measures of population trends. Shown here are four example population trend models, based on eBird data between 2010-2018, for Noisy Miner (top left), Hardhead (top right), Masked Lapwing (bottom left), and Crested Pigeon (bottom right), in the Greater Sydney Region, Australia.}
\label{fig1}
\end{figure}

The number of citizen science projects which are focused on ecological and environmental monitoring is increasing \cite{pocock2017diversity, theobald2015global}, highlighting the potential that citizen science holds for the future of ecology, conservation, and natural resource management \cite{pocock2018vision, silvertown2009new, soroye2018opportunistic, mckinley2017citizen}. But a major obstacle in the future use of citizen science data remains understanding how to best extract information from 'noisy' citizen science datasets \cite{parrish2018exposing}. As mentioned, this noise from citizen science \cite{bird2014statistical} can sometimes be alleviated using 'big-data' statistical approaches \cite{kelling2015taking}, but this is most applicable for data originating from large, successful citizen science projects --- with lots of data. And even with 'big data', this is not always possible. But what about projects that are just beginning? Or projects focusing on taxa that are less popular with the general public \cite{mair2016explaining, ward2014understanding}? Are there optimal strategies for sampling in space and time for estimating biodiversity trends?

Here, we investigate these questions with a specific objective: assess how spatial and temporal sampling influences trend detection of biodiversity. Our approach is dynamic: we are interested in the parameters that influence the marginal value of a given citizen science observation in both time and space. First, we summarize the parameters which we hypothesized would influence the marginal value of a citizen science observation.

@@

\begin{itemize}
  \item \textbf{Site sampling}: if a site has been previously sampled or not. We predicted that unsampled sites would be marginally more valuable than previously sampled sites.
  \item \textbf{Median sampling interval}: the median of the distribution of waiting times between samples at a site. We predicted that the median sampling interval would be positively associated with the value of a citizen science observation; i.e., observations from sites with high median waiting times would be more valuable than observations from sites with low median waiting times.
  \item \textbf{Days since last sample}: the number of days between samples at a site. We predicted that the number of days since the last sample would be positively associated with the value of a citizen science observation.
  \item \textbf{Distance to the nearest sampled site}: the distance between the site in question and the nearest sampled site. We predicted that the distance to the nearest sampled site would be positively associated with the value of a citizen science observation.
  \item \textbf{Nearest neighbor sampling interval}: the median sampling interval of the nearest neighbor. We predicted that the nearest-neighbor sampling interval would positively influence the value of an observation, whereby well-sampled areas (i.e., multiple sites near each other with low median sampling intervals) would have lower value.
  \item \textbf{Number of unique days sampled}: the total number of unique days sampled for a given site. We predicted that the total number of unique days would be positively associated with the value of an observation, whereby sites with many observations would have additional value given the long-term data originating from them.
\end{itemize}

We test our predictions using a popular citizen science project --- eBird \cite{sullivan2009ebird} --- and > 5 million bird observations from the Greater Sydney Region, NSW, Australia. We fit presence/absence population trend models to 235 species throughout this region, based on data from 2010 to 2018 (Fig. 1), and calculate statistical leverage for every citizen science sampling event, for each species. The leverage is a measure of how 'valuable' a given observation is to the population trend model for each species. Each sampling event is then assigned a cumulative measure of leverage --- i.e., marginal value. We then test our predictions above using linear models, assessing the influence each of the parameters has on the trend. To assess the robustness of our approach, we investigated our predictions at four different site scales, represented by grids of 5, 10, 25, and 50 km$^{2}$.

Identifying the contribution of each parameter to marginal value, we use the parameterized model results to predict the expected leverage (i.e., marginal value) for a given site in space and time, producing a dynamic map of expected marginal value of citizen science sampling for every day in 2018.

\section*{Results and Discussion}
\subsection*{Tests of predictions}
We found weak evidence that visiting an unsampled site was marginally more valuable than visiting an already sampled site. There was no statistical significance for the 5 km (p=0.669), and 10 km (p=0.093) site sizes, but there was for the 25 km (p=0.035) site size. At the 50 km site size, this test was not possible because all sites had been sampled. These results suggests that stratified sampling --- an approach which aims for equal sampling among sites --- \cite{baillie2008toward, longino1997biodiversity} is not the most applicable approach for detecting trends using citizen science data. In other words, citizen scientists are already sufficiently sampling biodiversity in space: they are appropriately identifying and sampling the 'hotspots' in space that should receive the most sampling attention.

@@

For those sampled sites, however, we found significant relationships between a number of our predicted parameters and the marginal value of a sampling event (Fig. 2). In general, our parameters were positively associated with the marginal value of a sampling event. The number of unique days sampled --- likely represented from known 'hotspots' identified by citizen scientists --- had the strongest, positive, effect size, and this was robust to the site size comparisons. The median sampling interval was also strongly associated with high marginal value samples, with an exception at the 25 km site size. Distance to the nearest sampled site and the nearest neighbor sampling interval influenced the marginal value of a sampling event less than the other parameters. Surprisingly, the number of days since the last sample, while positively associated, had less influence than the other parameters. The fact that days since last sample had a lower effect size than both the median sampling interval and the number of unique days sampled, highlights the value of `revisiting' a site (i.e., 'hotspot') in order to extract the maximum amount of information.

Together, these results suggest that improvements can be made in biodiversity sampling for trend detection. The `history' of a site is particularly important while considering whether to sample that site: the number of unique days sampled was the strongest predictor, suggesting that observations from sites with a long time-series are marginally more valuable. A secondary goal should be to decrease (i.e., left-shift the distribution) the median sampling interval of the sites. The effect of number of days sampled was strongest at the largest site size, suggesting that given all sites are sampled with some regularity --- as they are with larger site sizes --- the importance of the `history' of a site is inflated.

We found generally consistent results, based on the site size chosen, highlighting the robustness of spatial scale in our approach. It is critical to track biodiversity trends at multiple spatial scales \cite{soberon2007assessing}, as biodiversity estimates sometimes change dependent on the spatial scale \cite{chase2013scale}. Our approach of relying on statistical leverage as a measure of the value of a citizen science observation can be used regardless of whether a citizen science project is global, regional, or carried out in the constraints of a local park. In any instance, the area can be gridded --- the size of a grid, however, would be proportional to the spatial scale of the study --- and the framework implemented, updated dynamically.

\subsection*{Applications of our predictions}
Providing dynamic feedback to citizen science participants has proved successful for many citizen science projects \cite{rowley2019frogid, wiggins2011conservation, xue2016avicaching}. This feedback is generally in the form of leaderboards, presenting the number of submissions or number of unique species someone has contributed \cite{wood2011ebird}. But leaderboards tend to focus on outputs, incentivizing finding rather than looking, leading to perverse outcomes related to the redundancies and gaps in citizen science data. We sought to develop an outcome-based incentive by using our fitted models to predict the expected value of a given citizen science observation, dynamically, for any given day (e.g., Fig. 3, \href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/dynamic_map.gif}{Video 2}). This approach required us to look backward first, using a model with all observations for 2018, based on statistical leverage calculated from 2010-2018, in order to look forward and predict the expected leverage for any given day. We envision a dynamic approach (\href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/dynamic_map.gif}{Video 2}) in future citizen science projects, which would ultimately guide participants to sites which \textit{should} be sampled on any given day --- or in a given week, month, or year. In this instance, leaderboards would move past numbers of species or submissions and could be derived based on a participant's cumulative value to the citizen science dataset, for biodiversity outcomes. Instead of participants preferentially chasing specific species, this approach would guide participants to the sites with the highest expected marginal value. For example, we imagine visitor centers across the world at national parks or urban greenspaces providing their visitors on any given day a localized map showing which trail someone should visit if they are interested in contributing the greatest value to that park's biodiversity knowledge, through citizen science. The global pull of ecotourism \cite{sharpley2006ecotourism} is increasing exponentially, creating the potential for people to contribute to local biodiversity knowledge in areas that are traditionally undersampled, and with this framework, the collective effort of citizen scientists can be maximized.

\begin{figure}[!hb]
\centering
\includegraphics[width=.8\linewidth]{example_maps.png}
\caption{A map of predicted expected marginal value for six different days in 2018, throughout the Greater Sydney Region, showing the highest valued sites that would optimize the collective knowledge on biodiversity trends throughout the Greater Sydney Region. This prediction step is dynamic: the predictions are updated as new observations are submitted to the citizen science database. Prediction updating can be done at any frequency (see updates for every day in 2018 (\href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/dynamic_map.gif}{Video 2})).  Expected value maps will need to be updated fast in parts of the world where sampling rates are high, but this can be done at a slower rate where sampling is less frequent.}
\label{fig3}
\end{figure}

We focused our framework on a specific statistical outcome: trend detection. Many other ecological outcomes arise from citizen science datasets, including species distribution models \cite{bradsworth2017species, van2013opportunistic}, phylogeographical research \cite{bahls2014new, drury2019continent}, invasive species detection \cite{pocock2017citizen, grason2018citizen}, or phenological research \cite{la2014role, supp2015citizen}. Each potential outcome will have different optimal sampling strategies in space and time, likely with nuanced trade-offs between outcomes. But these different outcomes can still be quantified in the same framework we introduce here. For example, an intended outcome of a species distribution model would likely place greater value on observations from unsampled sites \cite{crawley2001scale} than for species trend detection. Our framework is robust, with the necessary information being a statistical model based on an intended ecological/management outcome.

\subsection*{Conclusions}
Since eBird's inception in 2002, citizen scientists have collectively contributed > 30 million effort hours. And this is only one citizen science project. Clearly, the cumulative effort put-forth by citizen scientists is immense, and arguably, citizen science will continue to shape the future of ecology and conservation --- as it has substantially for the past couple of centuries \cite{silvertown2009new} --- with an increasingly critical role in monitoring of biodiversity\cite{mckinley2017citizen, pocock2018vision}. But we need to look towards the future. Are there mechanisms we can put in place now which will increase our collective knowledge gleaned from citizen science datasets for biodiversity in the future? We highlight general rules which could help guide citizen science participants to better sampling in space and time: the number of unique days sampled and the largest median sampling intervals both positively correlate with the marginal value of a citizen science observation. Moreover, we demonstrate a framework which citizen science projects can implement to better optimize their sampling designs, which can be tailed to specific citizen science project goals.
